---
title: "STA 225 2.0 Design and Analysis of Experiments"
subtitle: "Lecture 1"
author: "Dr Thiyanga S. Talagala"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
      - xaringan-themer.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#081d58",
  secondary_color = "#FF961C",
 inverse_header_color = "#FFFFFF",
 title_slide_text_color = "#edf8b1",
 link_color =  "#41b6c4"
)
#style_solarized_light(text_font_google   = google_font("Josefin Sans", "400", "400i", "800i", "800"))
#style_mono_light(
#  base_color = "#1c5253",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Josefin Sans", "400", "400i", "800i", "800"),
#  code_font_google   = google_font("Fira Mono")
#)
```

<style>

.center2 {
  margin: 0;
  position: absolute;
  top: 50%;
  left: 50%;
  -ms-transform: translate(-50%, -50%);
  transform: translate(-50%, -50%);
}

</style>

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
}
</style>


class: center, middle

---

## Example 2

**Depression, anxiety, and stress among Sri Lankan undergraduates in times of Covid-19 lockdown**

For the purpose of data collection, an online cross-sectional survey was carried out from May 26 to June 27, 2020, utilizing a Google adapted preformed questionnaire. The form was shared with a short overview and justification through Facebook, Twitter, Facebook messenger, Viber, and What’s App. 

