---
title: "STA 225 2.0 Design and Analysis of Experiments"
subtitle: "Lecture 3"
author: "Dr Thiyanga S. Talagala"
date: "2021-10-28"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
      - xaringan-themer.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#081d58",
  secondary_color = "#FF961C",
 inverse_header_color = "#FFFFFF",
 title_slide_text_color = "#edf8b1",
 link_color =  "#41b6c4"
)
#style_solarized_light(text_font_google   = google_font("Josefin Sans", "400", "400i", "800i", "800"))
#style_mono_light(
#  base_color = "#1c5253",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Josefin Sans", "400", "400i", "800i", "800"),
#  code_font_google   = google_font("Fira Mono")
#)
```

<style>

.center2 {
  margin: 0;
  position: absolute;
  top: 50%;
  left: 50%;
  -ms-transform: translate(-50%, -50%);
  transform: translate(-50%, -50%);
}

</style>

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
}
</style>
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```



## Completely Randomized Design (CRD)

- The simplest experimental design - CRD

- The treatments are assigned completely at random so that each experimental unit has the same chance of receiving every treatment.

- Appropirate only for experiments with **homogeneous** experimental units

---

## Example

**Objective of the study:** Effect of fertilizer dose on the yield of chille

--

**Experimental design:** CRD

--

**Independent variable:** Fertilizer doses 

--

**Treatment:** Different doses of fertilizer (A, B, C, D, E)

--

**Dependent variables:** Number of chilles after one month

--

**Number of replicates:** 5

--

**Number of experimental units:** $\sum_{i=1}^{t}r_i = 25$, where $t$ is the number of treatments ( $t = 1, 2, 3, 4, 5$ ) and $r_i$ number of replicates under treatment $i$


---

## Step 1: Assign treatments at random



---

## Step 2: Measure the response variable


---
## Step 3: Organize data

```{r, comment=NA, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
A <- c(8, 8, 15, 11, 7)
B <- c(11, 17, 13, 18, 17)
C <- c(14, 16, 18, 20, 20)
D <- c(20, 24, 22, 19, 23)
E <- c(8, 10, 12, 15, 12)
df <- data.frame(A=A, B=B, C=C, D=D, E=E)
df %>% knitr::kable(format = "html", table.attr = "style='width:50%;'",  align = "c") %>% 
  kableExtra::kable_styling() %>% row_spec(1:5, background="green")
```

---


## Step 4: Exploratory data analysis

.pull-left[

```{r, comment=NA, echo=FALSE}
df <- df %>% pivot_longer(1:5, "Treatment", "value")
ggplot(df, aes(x = factor(1), y = value)) +
  geom_boxplot(width = 0.4, fill = "white") +
  geom_point(aes(color = Treatment)) + scale_color_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e"))  
```

]

--

.pull-right[

```{r, comment=NA, echo=FALSE}
ggplot(df, aes(x = factor(1), y = value)) +
  geom_boxplot(width = 0.4, fill = "white") +
  geom_jitter(aes(color = Treatment), width = 0.2, height = 0.00) + scale_color_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e"))  
```

]




---

## Step 4: Exploratory data analysis

.pull-left[

```{r, comment=NA, echo=FALSE}
#df <- df %>% pivot_longer(1:5, "Treatment", "Value")
ggplot(data=df, aes(x=Treatment, y=value, col=Treatment)) +  geom_point() + scale_color_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) 
```

]


.pull-right[

```{r, comment=NA, echo=FALSE}
ggplot(data=df, aes(x=Treatment, y=value, col=Treatment)) + geom_boxplot(aes(fill=Treatment), alpha=0.5) +   geom_point(colour="black", alpha=0.5) + scale_color_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) +  scale_fill_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) 
```


]

---

.pull-left[

```{r, comment=NA, message=FALSE, warning=FALSE}
library(tidyverse)
A <- c(8, 8, 15, 11, 7)
B <- c(11, 17, 13, 18, 17)
C <- c(14, 16, 18, 20, 20)
D <- c(20, 24, 22, 19, 23)
E <- c(8, 10, 12, 15, 12)
df <- data.frame(A=A, B=B, C=C, D=D, E=E)
df <- df %>% pivot_longer(1:5, "Treatment", "value")
df$Treatment <- as.factor(df$Treatment)
df %>% group_by(Treatment) %>% summarise(
    mean=mean(value),
    sd=sd(value),
    median=median(value),
    min=min(value),
    max=max(value))
```

]

.pull-right[

```{r, comment=NA, echo=FALSE}
ggplot(data=df, aes(x=Treatment, y=value, col=Treatment)) + geom_boxplot(aes(fill=Treatment), alpha=0.5) +   geom_point(colour="black", alpha=0.5) + scale_color_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) +  scale_fill_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) 
```


]

---

## Sources of Variation

- Source of variation due to **treatments**


- Source of variation due to **experimental error**

## Partition of total variability

<style>
div.blue pre { background-color:lightblue; }
div.blue pre.r { background-color:red; }
</style>

<div class = "red">
$$\textbf{Variability}_\text{total} = \textbf{Variability}_\text{treatment} + \textbf{Variability}_\text{experimental error}$$
</div>


---

## Notations

| Treatment (level) 	|  R1 	|  R2	|  ...	| Rn 	| Totals 	|  Averages	|
|:---:	| :-----:	| :---:	| :---:	| :---:	| :---:	| :---:	|
|  1	| $y_{11}$ 	|  $y_{12}$	|  ...	| $y_{1n}$	|  $y_{1.}$	| $\bar{y}_{1.}$ 	|
|  2	| $y_{21}$ 	| $y_{22}$ 	| ... 	| $y_{2n}$ 	| $y_{2.}$ 	| $\bar{y}_{2.}$	|
| . 	|  	|  	|  	|  	|  	|  	|
| . 	|  	|  	|  	|  	|  	|  	|
| . 	|  	|  	|  	|  	|  	|  	|
|  a	| $y_{a1}$ 	| $y_{a2}$ 	|  ...	| $y_{an}$ 	| $y_{a.}$ 	| $\bar{y}_{a.}$ 	|
|  	|  	| 	|  	| 	| $y_{..}$ 	| $\bar{y}_{..}$ 	|


---


| Treatment (level) 	|  R1 	|  R2	| R3| R4	| R5 	| Totals 	|  Averages  	|
|:---:	| :-----:	| :---:	| :---:	| :---:	| :---:	| :---:	|
|  A	| $y_{11}$ <br> 8 	| $y_{12}$ <br> 8	| $y_{13}$ <br> 15 | $y_{14}$ <br> 11	| $y_{15}$ <br> 7	|  $y_{1.}$  <br> 49	| $\bar{y}_{1.}$ <br> 9.8	|
|  B	| $y_{21}$ <br> 11 	| $y_{22}$ <br> 17	| $y_{23}$ <br> 13 | $y_{24}$ <br> 18	| $y_{25}$ <br> 17	|  $y_{2.}$ <br>	76| $\bar{y}_{2.}$ <br> 15.2 	|
|  C	| $y_{31}$ <br> 14 	| $y_{32}$ <br> 16	| $y_{33}$ <br> 18 | $y_{34}$ <br> 20	| $y_{35}$ <br> 20	|  $y_{3.}$ <br>	88| $\bar{y}_{3.}$ <br> 17.6 	|
|  D	| $y_{41}$ <br> 20 	| $y_{42}$ <br> 24	| $y_{43}$ <br> 22 | $y_{44}$ <br> 19	| $y_{45}$ <br> 23	|  $y_{4.}$ <br> 108	| $\bar{y}_{4.}$ <br> 	21.6|
|  E	| $y_{51}$ <br> 8 	| $y_{52}$ <br> 10	| $y_{53}$ <br> 12 | $y_{54}$ <br> 15	| $y_{55}$ <br> 12	|  $y_{5.}$ <br>	57| $\bar{y}_{5.}$ <br> 11.4 	|
|  	|  	| 	| | 	| 	|  $y_{..} = 378$| $\frac{378}{25} = \frac{75.6}{5} = 15.12$ 	|


---

## Notations (cont.)

$a$ - number of different treatments

$y_{ij}$ - $j$th observation taken under factor level or treatment $i$

$n$ - observations under the $i$th treatment

$y_{i.}$ - total of the observations under the $i$th treatment

$\bar{y}_{i.}$ - average of the observations under the $i$th treatment

$y_{i.} = \sum_{j=1}^n y_{ij},$ 

$\bar{y}_{i.} = \frac{y_{i.}}{n},$ $i = 1, 2, ..., a$


---

## Notations (cont.)

$y_{..}$ - grand total of all the observations

$\bar{y}_{..}$ - grand average of all the observations



$y_{..} = \sum_{i=1}^a\sum_{j=1}^n y_{ij}$


$\bar{y}_{..} = \frac{y_{..}}{N},$ where $N = an$



Note: "dot" subscript notation implies summation over the subscript that it replaces.

---
class: inverse, center, middle

#**one-way or single-factor analysis of variance**

---

## Model for the data

We can describe the observations come from an experiment using a model. There are two ways to write the models

**Method 1: means model**

$$Y_{ij}= \mu_i + \epsilon_{ij}
\begin{cases}
     i=1, 2, ..., a \\
    j=1, 2, ..., n           
\end{cases}$$

Where $Y_{ij}$ is the $ij$th observation, $\mu_i,$ is the mean of the $i$th factor level or treatment, and $\epsilon_{ij}$ is the **random error** component that includes other variabilities such as background noise, differences between experimental units, variability arising from uncontrolled factors, etc.

We assume errors have mean zero, hence, $E(Y_{ij}) = \mu_{i}$.

---

**Method 2: Effects model (an alternative way to write the mean model)**

$$\mu_i = \mu + \tau_i, \text{  } i=1, 2, ..., a$$

Hence the following equation can be written as

$$Y_{ij}= \mu_i + \epsilon_{ij}
\begin{cases}
     i=1, 2, ..., a \\
    j=1, 2, ..., n           
\end{cases}$$

$$Y_{ij}= \mu + \tau_i + \epsilon_{ij}
\begin{cases}
     i=1, 2, ..., a \\
    j=1, 2, ..., n           
\end{cases}$$

where $\mu$, is common to all treatment and we called it **overall mean**. The parameter $\tau_i$ is unique to the $i$th
treatment called the $i$**th treatment effect.** 

---

## Model assumptions

The model errors are assumed to be normally and independently distributed random variables with mean zero and variance $\sigma^2$. The variance $\sigma^2$ is assumed to be constant for all levels of the factor. 

This implies that the observations

$$y_{ij} \sim N(\mu+\tau_i, \sigma^2)$$

and that the observations are mutually independent.

---

## In-class


$$Y_{ij}= \mu + \tau_i + \epsilon_{ij}
\begin{cases}
     i=1, 2, ..., a \\
    j=1, 2, ..., n           
\end{cases}$$

This is called the **random effects model** or **components of variance model**.

---

## Our goal

- Test appropriate hypotheses about the treatment means

- Estimate treatment means

---

## Hypotheses

We want to test the equality of the $a$ treatment meann, that is, $E(Y_{ij}) = \mu + \tau_i = \mu_i, i=1, 2, ..., 1.$



In terms of $\mu_i$

$$H_0: \mu_1 = \mu_2=...=\mu_a$$
$$H_1: \mu_i \neq \mu_j \text{ for at least one pair (i, j)}$$

In terms of $\tau_i$

$$H_0: \tau_1 = \tau_2=...=\tau_a=0$$
$$H_1: \tau_i \neq 0 \text{ for at least one }i$$
---

### Analysis of Variance (ANOVA)


$$\textbf{Variability}_\text{total} = \textbf{Variability}_\text{treatment} + \textbf{Variability}_\text{experimental error}$$


ANOVA is derived from decomposing the total sum of squares (partitioning of total variability into its component). 

The total sum of squares ( $SS_T$ )

$$SS_T = \sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2.$$

$SS_T$ measures the overall variability in the data.

Number of degrees of freedom = $an-1 = N-1$

---

## Decomposition of the Total Sum of Squares

$$\sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2 = \sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..} + \bar{y}_{i.} - \bar{y}_{i.})^2$$

$$\sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2 = \sum_{i=1}^a\sum_{j=1}^n [(\bar{y}_{i.} - \bar{y}_{..} ) + (y_{ij} - \bar{y}_{i.})]^2$$

$\sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2 = n \sum_{i=1}^a(\bar{y}_{i.} - \bar{y}_{..})^2 + \sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{i.})^2$
$\text{            }+ 2\sum_{i=1}^a\sum_{j=1}^n(\bar{y}_{i.} - \bar{y}_{..})(y_{ij}-\bar{y}_{i.})$

The cross product term is zero. Hence,

$\sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2 = n \sum_{i=1}^a(\bar{y}_{i.} - \bar{y}_{..})^2 + \sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{i.})^2$


---

$\sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{..})^2 = n \sum_{i=1}^a(\bar{y}_{i.} - \bar{y}_{..})^2 + \sum_{i=1}^a\sum_{j=1}^n (y_{ij} - \bar{y}_{i.})^2$

$$SS_T = SS_{Treatments} + SS_{E}$$


$SS_{Treatments}$ is called the sum of squares due to treatments (i.e., between treatments)

$SS_{E}$ is called the sum of squares due to error (i.e., within treatments)

---

## ANOVA Table

Source of variation |  Sum of squares (SS) | DF| Mean Square (MS) | F| p-value |
---:---:---:---:---:---|
Between treatments |  $SS_{Treatments}$ | $a-1$ | $MS_{Treatments}$ | $F_0=\frac{MS_{Treatments}}{MS_E}$ | $P(F \geq F_0)$|
Error (within treatments) | $SS_E$ | $N-a$ | $MS_{E}$ | |
Total |  $SS_T$ | $N-1$| | | |


---

$$SS_E = SS_T - SS_{Treatments}$$

$$MS_{Treatments} = \frac{SS_{Treatments}}{a-1}$$

$$MS_{E} = \frac{SS_{E}}{N-a}$$

$$F_0 = \frac{MS_{Treatments}}{MS_E}$$

If the null hypothesis of no difference in treatment means is true, the ratio

$$F_0 = \frac{MS_{Treatments}}{MS_E}$$

follows a $F$ distribution with $a-1$ and $N-a$ degrees of freedom.

If $F_0 > F_{\alpha, a-1, N-a }$, we can conclude that there are differences in the treatment means.


---

.pull-left[

```{r, comment=NA, message=FALSE, echo=FALSE}
df$Treatment <- as.factor(df$Treatment)
df %>% group_by(Treatment) %>% summarise(
    mean=mean(value),
    sd=sd(value),
    median=median(value),
    min=min(value),
    max=max(value))
```

What is your guess? Reject $H_0$ or Do not reject $H_0$?

$$H_0: \mu_1 = \mu_2=...=\mu_5$$
$$H_1: \mu_i \neq \mu_j \text{ for at least one pair (i, j)}$$

]

.pull-right[

```{r, comment=NA, echo=FALSE}
ggplot(data=df, aes(x=Treatment, y=value, col=Treatment)) + geom_boxplot(aes(fill=Treatment), alpha=0.5) +   geom_point(colour="black", alpha=0.5) + scale_color_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) +  scale_fill_manual(values=c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")) 
```


]

---

# ANOVA

```{r, comment=NA}
one.way <- aov(value~ Treatment, data = df)
summary(one.way)
```

