---
title: "STA 225 2.0 Design and Analysis of Experiments"
subtitle: "Lecture 3"
author: "Dr Thiyanga S. Talagala"
date: "2021-10-28"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
      - xaringan-themer.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#081d58",
  secondary_color = "#FF961C",
 inverse_header_color = "#FFFFFF",
 title_slide_text_color = "#edf8b1",
 link_color =  "#41b6c4"
)
#style_solarized_light(text_font_google   = google_font("Josefin Sans", "400", "400i", "800i", "800"))
#style_mono_light(
#  base_color = "#1c5253",
#  header_font_google = google_font("Josefin Sans"),
#  text_font_google   = google_font("Josefin Sans", "400", "400i", "800i", "800"),
#  code_font_google   = google_font("Fira Mono")
#)
```

<style>

.center2 {
  margin: 0;
  position: absolute;
  top: 50%;
  left: 50%;
  -ms-transform: translate(-50%, -50%);
  transform: translate(-50%, -50%);
}

</style>

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
}
</style>
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```



## Completely Randomized Design (CRD)

- The simplest experimental design - CRD

- The treatments are assigned completely at random so that each experimental unit has the same chance of receiving every treatment.

- Appropirate only for experiments with **homogeneous** experimental units

---

## Example

**Objective of the study:** Effect of fertilizer dose on the yield of chille

--

**Experimental design:** CRD

--

**Independent variable:** Fertilizer doses 

--

**Treatment:** Different doses of fertilizer (A, B, C, D, E)

--

**Dependent variables:** Number of chilles after one month

--

**Number of replicates:** 5

--

**Number of experimental units:** $\sum_{i=1}^{t}r_i = 25$, where $t$ is the number of treatments ( $t = 1, 2, 3, 4, 5$ ) and $r_i$ number of replicates under treatment $i$


---

Step 1: Assign treatments at random



---

Step 2: Measure the response variable


---
Step 3: Organize data

```{r, comment=NA, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
A <- c(8, 8, 15, 11, 7)
B <- c(11, 17, 13, 18, 17)
C <- c(14, 16, 18, 20, 20)
D <- c(20, 24, 22, 19, 23)
E <- c(8, 10, 12, 15, 12)
df <- data.frame(A=A, B=B, C=C, D=D, E=E)
df %>% knitr::kable(format = "html", table.attr = "style='width:50%;'",  align = "c") %>% 
  kableExtra::kable_styling() %>% row_spec(1:5, background="green")
```

---

Step 4: Exploratory data analysis

.pull-left[

```{r, comment=NA, eval=F}
df <- df %>% pivot_longer()
```

]

.pull-right[


]

---

## Sources of Variation

- Source of variation due to **treatments**


- Source of variation due to **experimental error**

## Partition of total variability

<style>
div.blue pre { background-color:lightblue; }
div.blue pre.r { background-color:red; }
</style>

<div class = "red">
$$\textbf{Variability}_\text{total} = \textbf{Variability}_\text{treatment} + \textbf{Variability}_\text{experimental error}$$
</div>


---

## Notations

| Treatment (level) 	|  R1 	|  R2	|  ...	| Rn 	| Totals 	|  Averages	|
|:---:	| :-----:	| :---:	| :---:	| :---:	| :---:	| :---:	|
|  1	| $y_{11}$ 	|  $y_{12}$	|  ...	| $y_{1n}$	|  $y_{1.}$	| $\bar{y}_{1.}$ 	|
|  2	| $y_{21}$ 	| $y_{22}$ 	| ... 	| $y_{2n}$ 	| $y_{2.}$ 	| $\bar{y}_{2.}$	|
| . 	|  	|  	|  	|  	|  	|  	|
| . 	|  	|  	|  	|  	|  	|  	|
| . 	|  	|  	|  	|  	|  	|  	|
|  a	| $y_{a1}$ 	| $y_{a2}$ 	|  ...	| $y_{an}$ 	| $y_{a.}$ 	| $\bar{y}_{a.}$ 	|
|  	|  	| 	|  	| 	| $y_{..}$ 	| $\bar{y}_{..}$ 	|


---


| Treatment <br> (level) 	|  R1 	|  R2	| R3| R4	| R5 	| Totals 	|  Averages	|
|:---:	| :-----:	| :---:	| :---:	| :---:	| :---:	| :---:	|
|  A	| $y_{11}$ <br> 8 	| $y_{12}$ <br> 8	| $y_{13}$ <br> 15 | $y_{14}$ <br> 11	| $y_{15}$ <br> 7	|  $y_{1.}$ <br>	| $\bar{y}_{1.}$ <br> 	|
|  B	| $y_{21}$ <br> 11 	| $y_{22}$ <br> 17	| $y_{23}$ <br> 13 | $y_{24}$ <br> 18	| $y_{25}$ <br> 17	|  $y_{2.}$ <br>	| $\bar{y}_{2.}$ <br> 	|
|  C	| $y_{31}$ <br> 14 	| $y_{32}$ <br> 16	| $y_{33}$ <br> 18 | $y_{34}$ <br> 20	| $y_{35}$ <br> 20	|  $y_{3.}$ <br>	| $\bar{y}_{3.}$ <br> 	|
|  D	| $y_{41}$ <br> 20 	| $y_{42}$ <br> 24	| $y_{43}$ <br> 22 | $y_{44}$ <br> 19	| $y_{45}$ <br> 23	|  $y_{4.}$ <br>	| $\bar{y}_{4.}$ <br> 	|
|  E	| $y_{51}$ <br> 8 	| $y_{52}$ <br> 10	| $y_{53}$ <br> 12 | $y_{54}$ <br> 15	| $y_{55}$ <br> 12	|  $y_{5.}$ <br>	| $\bar{y}_{5.}$ <br> 	|


---


